---
title: "Model accuracy report"
subtitle: "Data Science Capstone — predictive evaluation"
author: "Aseem Anand"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: readable
    df_print: kable
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 5
)
```

This report continues **`capstone_milestone_report.Rmd`**: English SwiftKey text (blogs, news, Twitter) from **`Coursera-SwiftKey.zip`** or extracted **`final/en_US/`**, using the **same token rule** as the milestone EDA (`extract_alpha_tokens`: lowercase words with letters and apostrophes). The milestone uses larger head samples for plots; here **`lines_per_source`** is smaller so training and Bachman-style evaluation knit quickly—raise it for stronger models.

A random **train/test split** at the **line** level feeds Katz **n-gram** tables (training) and **test-set n-gram type frequencies** (evaluation instances).

The **evaluation protocol** follows the rank-based accuracy summary described in the [Model Accuracy Report](https://shelbybachman.github.io/data-science-capstone/reports/Model-Accuracy-Report.html): sample high- and low-frequency 3-grams and 4-grams from the test corpus, ask the model for the **50** most likely next tokens given the typed prefix, and convert ranks into a smooth accuracy score.

```{r deps}
if (!requireNamespace("data.table", quietly = TRUE)) {
  install.packages("data.table", repos = "https://cloud.r-project.org")
}
library(dplyr)
library(ggplot2)
library(knitr)

source("R/milestone_paths.R", encoding = "UTF-8")
source("R/milestone_load_samples.R", encoding = "UTF-8")
source("R/source_lm_pipeline.R", encoding = "UTF-8")

swiftkey <- resolve_swiftkey_en()
assert_swiftkey_data_available(swiftkey)

# Head sample per source file — larger samples yield richer test n-gram bins (slower knit).
lines_per_source <- 2200L
train_frac <- 0.80
katz_discount <- 0.75
n_top_predictions <- 50L
n_samples_per_bin <- 25L

raw_lines <- assemble_head_samples(swiftkey, lines_per_source)$text
sp <- train_test_tokens_from_lines(raw_lines, train_frac = train_frac, seed = 42L)
```

Training incorporates **`r round(train_frac * 100)`%** of the sampled lines (chosen at random); testing uses the remainder so training lines never appear as raw lines in the test pool. From **tokenized test lines only**, every distinct trigram (`start` = two words, `end` = third) and quadgram (`start` = three words, `end` = fourth) receives a **frequency** count.

**High-frequency test types** are those with frequency **> 50**; **low-frequency** types are those with frequency **> 10** and **< 25** (matching the cited report’s bins). Up to **`r n_samples_per_bin`** instances are sampled from each bin for both 3-grams and 4-grams. The **same** list of prefixes and target words is reused for both trained models below so comparisons are paired.

## Models compared

1. **Full table:** retain bigrams, trigrams, and quadgrams with count ≥ 1 (plus unigrams).
2. **App-style pruning:** retain higher-order n-grams only when count **> 2** (i.e. stored types with frequency ≥ 3), analogous to the “training set for app” contrast in the reference report.

Both models combine **multiple n-gram orders** with **Katz backoff**: at each order, continuation probabilities use **absolute discounting** with **`D = ` `r katz_discount`** subtracted from each nonzero joint count; the remaining probability mass is spread according to the **shortened history** (drop the earliest context word), down to **unigrams**. Candidate words are ranked by these backoff probabilities over the model’s candidate set; the top **`r n_top_predictions`** are returned for scoring.

```{r train}
lm_full <- build_ngram_lm(
  sp$train,
  min_freq_uni = 1L,
  min_freq_bi = 1L,
  min_freq_tri = 1L,
  min_freq_quad = 1L
)

lm_app <- build_ngram_lm(
  sp$train,
  min_freq_uni = 1L,
  min_freq_bi = 3L,
  min_freq_tri = 3L,
  min_freq_quad = 3L
)

kable(
  tibble(
    model = c("Full (all retained types)", "Pruned (count > 2)"),
    ngram_rows = c(lm_n_parameters(lm_full), lm_n_parameters(lm_app))
  ),
  caption = "Approximate model size (rows stored across uni–quad tables)"
)
```

```{r instances}
instances <- build_bachman_eval_instances(
  sp$test,
  n_samples = n_samples_per_bin,
  seed = 17L
)

if (nrow(instances) == 0L) {
  stop(
    "No evaluation instances — increase lines_per_source or relax frequency cuts in sample_ngram_eval_pairs().",
    call. = FALSE
  )
}

acc_full <- evaluate_lm_on_instances(
  instances,
  lm_full,
  katz_discount = katz_discount,
  n_to_return = n_top_predictions
)
acc_app <- evaluate_lm_on_instances(
  instances,
  lm_app,
  katz_discount = katz_discount,
  n_to_return = n_top_predictions
)

accuracy_long <- dplyr::bind_rows(
  dplyr::mutate(acc_full, training = "All retained n-grams"),
  dplyr::mutate(acc_app, training = "N-grams with count > 2")
)

accuracy_long$ngram <- factor(accuracy_long$ngram, levels = c("3-gram", "4-gram"))
accuracy_long$training <- factor(
  accuracy_long$training,
  levels = c("All retained n-grams", "N-grams with count > 2")
)
```

## Accuracy definition (reference methodology)

For each evaluation pair `(start, end)` the model returns the top **`r n_top_predictions`** candidates with columns `end` (candidate word), `score`, and `rank`.

- If the true word **`end`** is ranked **first**, accuracy = **1**.
- If it appears at rank **`r`** (2–`r n_top_predictions`), accuracy = **`(K - r + 1) / K`** with **`K = ` `r n_top_predictions`**.
- If it does **not** appear among the returned candidates, accuracy = **0**.

```{r accuracy-summary}
summ <- accuracy_long %>%
  group_by(training, ngram, freq_level) %>%
  summarise(mean_accuracy = mean(accuracy), .groups = "drop")
kable(summ, digits = 3, caption = "Mean Bachman-style accuracy by facet")
```

## Accuracy versus test-set type frequency

```{r accuracy-plot}
ggplot(
  dplyr::arrange(accuracy_long, ngram, freq_test),
  aes(x = freq_test, y = accuracy, colour = ngram, group = ngram)
) +
  geom_point(alpha = 0.85, na.rm = TRUE) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.7, formula = y ~ x) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~training, nrow = 1L) +
  labs(
    x = "Frequency of n-gram type in held-out test text",
    y = "Accuracy",
    colour = "",
    title = "Next-word accuracy vs. frequency of test n-gram types"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Computational effort

Timing covers **identical** `(start, end)` pairs for each model: wall-clock seconds for all `predict_next_word` calls.

```{r timing}
tm_full <- time_prediction_pairs(
  instances$start,
  instances$end,
  lm_full,
  katz_discount = katz_discount,
  n_to_return = n_top_predictions
)
tm_app <- time_prediction_pairs(
  instances$start,
  instances$end,
  lm_app,
  katz_discount = katz_discount,
  n_to_return = n_top_predictions
)

kable(
  tibble(
    model = c("Full", "Pruned"),
    seconds = c(tm_full$elapsed_sec, tm_app$elapsed_sec),
    predictions = c(tm_full$n, tm_app$n)
  ),
  digits = 3,
  caption = "Prediction timing on the fixed evaluation set"
)
```

---

**Interpretation:** pruning removes rare contexts so tables shrink (`lm_n_parameters`) and timings usually improve for Shiny-like workloads; accuracy versus rare test bins (`freq_level == "low frequency"`) is the usual casualty unless backoff bridges gaps effectively.
